Cascaded Approximation of Gaussian Blur

### 2015-05-19 article.tex
Add fourth section

\title{Cascaded Approximation of Gaussian Blur}
\author{Vabishchevich Nikolay}
\date{May 15, 2015}
\maketitle

@@

For last task of approximating central gaussian shape I've decided to use generic 9-tap filter as
the main workhorse. Under that name I imply even symmetric conservative convolutional filter of the
following form
\begin{equation}
    K_k = \left\{
        \begin{array}{ll}
            1 - \sum_{i=1}^4 C_i,& k = 0;\\
            \frac12 C_i,& k = m_i;\\
            0,& \text{otherwise};
        \end{array}
    \right.
\end{equation}
where $m_i$, $i=1, 2, 3, 4$, are sampling positions ($0 < m_1 < m_2 < m_3 < m_4$) and $C_i$ are
coefficients. Fourier transform of this filter kernel is
\begin{equation}
    K(\varphi) = 1 - \sum_{i=1}^4 C_i[1 - \cos(m_i\varphi)].
\end{equation}
Further research shows that this filter isn't enough as central filter so I combine it with
additional stage of even binomial filter $\hat B^{2l}$ ($B^0$ mean ``do nothing''). Therefore
complete approximation of gaussian blur take the form of
\begin{equation}
    \hat G_{\text{apx}} = \hat Q^n\hat K\hat B^{2l}\hat P^n.
\end{equation}
Disregarding frequency crosstalk, diagonal part of that operator in Fourier space is
\begin{equation}
    G_{\text{apx}}(\varphi) = \left\{1 - \sum_{i=1}^4 C_i[1 - \cos(2^nm_i\varphi)]\right\}
        \cos^{2l}(2^{n-1}\varphi) \prod_{k=0}^{n-1} \cos^{10}(2^{k-1}\varphi).
\end{equation}
Outside of the region $[-\pi/2^n, \pi/2^n]$ that should be already near zero, so we focus only on
center
\begin{equation}
    G_{\text{apx}}\left(\frac\varphi{2^n}\right) =
        \left\{1 - \sum_{i=1}^4 C_i[1 - \cos(m_i\varphi)]\right\}
        \cos^{2l}\frac\varphi2 \prod_{k=1}^n \cos^{10}\frac\varphi{2^{k+1}},
\end{equation}
where that expression should be near
\begin{equation}
    G\left(\frac\varphi{2^n}\right) = \exp\left\{-\frac{\sigma^2\varphi^2}{2\cdot4^n}\right\}.
\end{equation}

One of the possible ways to find good kernel coefficients $C_i$ is to solve least squares problem.
Denote
\begin{equation}
    p(\varphi) \equiv \cos^{2l}\frac\varphi2 \prod_{k=1}^n \cos^{10}\frac\varphi{2^{k+1}}
\end{equation}
and
\begin{equation}
    g(\varphi) \equiv \exp\left\{-\frac{\sigma^2\varphi^2}{2\cdot4^n}\right\},
\end{equation}
then the problem is to minimize
\begin{equation*}
    \lambda = \int_{-\pi}^\pi \left[G\left(\frac\varphi{2^n}\right) -
        G_{\text{apx}}\left(\frac\varphi{2^n}\right)\right]^2 w(\varphi) d\varphi =
\end{equation*}
\begin{equation*}
    \int_{-\pi}^\pi \left[g(\varphi) -
        \left\{1 - \sum_{i=1}^4 C_i[1 - \cos(m_i\varphi)]\right\}p(\varphi)\right]^2
        w(\varphi) d\varphi =
\end{equation*}
\begin{equation}
    \int_{-\pi}^\pi \left[g(\varphi) - p(\varphi) +
        p(\varphi)\sum_{i=1}^4 C_i[1 - \cos(m_i\varphi)]\right]^2 w(\varphi) d\varphi,
\end{equation}
where $w(\varphi) \approx 1$ is additional weight. As usual, calculating partial derivatives
\begin{equation*}
    0 = \frac12\frac{\partial\lambda}{\partial C_k} =
    \int_{-\pi}^\pi \left[g(\varphi) - p(\varphi) +
        p(\varphi)\sum_{i=1}^4 C_i[1 - \cos(m_i\varphi)]\right]
        p(\varphi)w(\varphi) [1 - \cos(m_k\varphi)] d\varphi =
\end{equation*}
\begin{equation}
    \int_{-\pi}^\pi [g(\varphi) - p(\varphi)]p(\varphi)w(\varphi) [1 - \cos(m_k\varphi)] d\varphi +
    \sum_{i=1}^4 C_i\int_{-\pi}^\pi p^2(\varphi)w(\varphi)
        [1 - \cos(m_i\varphi)][1 - \cos(m_k\varphi)] d\varphi.
\end{equation}
After defining
\begin{equation}
    U_k \equiv \frac1{2\pi}\int_{-\pi}^\pi p^2(\varphi)w(\varphi)e^{-ip\varphi}d\varphi
\end{equation}
and
\begin{equation}
    V_k \equiv \frac1{2\pi}\int_{-\pi}^\pi g(\varphi)p(\varphi)w(\varphi)e^{-ip\varphi}d\varphi
\end{equation}
we can get
\begin{equation}
    0 = \frac1{2\pi}\frac{\partial\lambda}{\partial C_k} = 2[V_0 - V_{m_k} - U_0 + U_{m_k}] +
        \sum_{i=1}^4 C_i[2U_0 - 2U_{m_i} - 2U_{m_k} + U_{m_i+m_k} + U_{|m_i-m_k|}].
\end{equation}
Resulting symmetric system of four linear equations
\begin{equation}
    \sum_{i=1}^4 C_i[2U_0 - 2U_{m_i} - 2U_{m_k} + U_{m_i+m_k} + U_{|m_i-m_k|}] =
        2[U_0 - U_{m_k} - V_0 + V_{m_k}],
\end{equation}
can be solved through standard methods of linear algebra.

@@

>>>>

\title{Cascaded Approximation of Gaussian Blur}
\author{Vabishchevich Nikolay}
\date{May 19, 2015}
\maketitle

@@

For last task of approximating central gaussian shape I've decided to use generic 9-tap filter as
the main workhorse. Under that name I imply even symmetric conservative convolutional filter of the
following form
\begin{equation}\label{main}
    K_k = \left\{
        \begin{array}{ll}
            1 - \sum_{i=1}^4 C_i,& k = 0;\\
            \frac12 C_i,& k = m_i;\\
            0,& \text{otherwise};
        \end{array}
    \right.
\end{equation}
where $m_i$, $i=1, 2, 3, 4$, are sampling positions ($0 < m_1 < m_2 < m_3 < m_4$) and $C_i$ are
coefficients. Fourier transform of this filter kernel is
\begin{equation}
    K(\varphi) = 1 - \sum_{i=1}^4 C_i[1 - \cos(m_i\varphi)].
\end{equation}
Further research shows that this filter isn't enough as central filter so I combine it with
additional stage of even binomial filter $\hat B^{2l}$ ($B^0$ mean ``do nothing''). Therefore
complete approximation of gaussian blur take the form of
\begin{equation}
    \hat G_{\text{apx}} = \hat Q^n\hat K\hat B^{2l}\hat P^n.
\end{equation}
Disregarding frequency crosstalk, diagonal part of that operator in Fourier space is
\begin{equation}
    G_{\text{apx}}(\varphi) = \left\{1 - \sum_{i=1}^4 C_i[1 - \cos(2^nm_i\varphi)]\right\}
        \cos^{2l}(2^{n-1}\varphi) \prod_{k=0}^{n-1} \cos^{10}(2^{k-1}\varphi).
\end{equation}
Outside of the region $[-\pi/2^n, \pi/2^n]$ that should be already near zero, so we focus only on
center
\begin{equation}
    G_{\text{apx}}\left(\frac\varphi{2^n}\right) =
        \left\{1 - \sum_{i=1}^4 C_i[1 - \cos(m_i\varphi)]\right\}
        \cos^{2l}\frac\varphi2 \prod_{k=1}^n \cos^{10}\frac\varphi{2^{k+1}},
\end{equation}
where that expression should be near
\begin{equation}
    G\left(\frac\varphi{2^n}\right) = \exp\left\{-\frac{\sigma^2\varphi^2}{2\cdot4^n}\right\}.
\end{equation}

One of the possible ways to find good kernel coefficients $C_i$ is to solve least squares problem.
Denote
\begin{equation}
    p(\varphi) \equiv \cos^{2l}\frac\varphi2 \prod_{k=1}^n \cos^{10}\frac\varphi{2^{k+1}}
\end{equation}
and
\begin{equation}
    g(\varphi) \equiv \exp\left\{-\frac{\sigma^2\varphi^2}{2\cdot4^n}\right\},
\end{equation}
then the problem is to minimize
\begin{equation*}
    \lambda = \int_{-\pi}^\pi \left[G\left(\frac\varphi{2^n}\right) -
        G_{\text{apx}}\left(\frac\varphi{2^n}\right)\right]^2 w(\varphi) d\varphi =
\end{equation*}
\begin{equation*}
    \int_{-\pi}^\pi \left[g(\varphi) -
        \left\{1 - \sum_{i=1}^4 C_i[1 - \cos(m_i\varphi)]\right\}p(\varphi)\right]^2
        w(\varphi) d\varphi =
\end{equation*}
\begin{equation}\label{lambda}
    \int_{-\pi}^\pi \left[g(\varphi) - p(\varphi) +
        p(\varphi)\sum_{i=1}^4 C_i[1 - \cos(m_i\varphi)]\right]^2 w(\varphi) d\varphi,
\end{equation}
where $w(\varphi) \approx 1$ is additional weight. As usual, calculating partial derivatives
\begin{equation*}
    0 = \frac12\frac{\partial\lambda}{\partial C_k} =
    \int_{-\pi}^\pi \left[g(\varphi) - p(\varphi) +
        p(\varphi)\sum_{i=1}^4 C_i[1 - \cos(m_i\varphi)]\right]
        p(\varphi)w(\varphi) [1 - \cos(m_k\varphi)] d\varphi =
\end{equation*}
\begin{equation}
    \int_{-\pi}^\pi [g(\varphi) - p(\varphi)]p(\varphi)w(\varphi) [1 - \cos(m_k\varphi)] d\varphi +
    \sum_{i=1}^4 C_i\int_{-\pi}^\pi p^2(\varphi)w(\varphi)
        [1 - \cos(m_i\varphi)][1 - \cos(m_k\varphi)] d\varphi.
\end{equation}
After defining
\begin{equation}
    U_k \equiv \frac1{2\pi}\int_{-\pi}^\pi p^2(\varphi)w(\varphi)e^{-ip\varphi}d\varphi
\end{equation}
and
\begin{equation}
    V_k \equiv \frac1{2\pi}\int_{-\pi}^\pi g(\varphi)p(\varphi)w(\varphi)e^{-ip\varphi}d\varphi
\end{equation}
we can get
\begin{equation}
    0 = \frac1{2\pi}\frac{\partial\lambda}{\partial C_k} = 2[V_0 - V_{m_k} - U_0 + U_{m_k}] +
        \sum_{i=1}^4 C_i[2U_0 - 2U_{m_i} - 2U_{m_k} + U_{m_i+m_k} + U_{|m_i-m_k|}].
\end{equation}
Resulting symmetric system of four linear equations
\begin{equation}
    \sum_{i=1}^4 C_i[2U_0 - 2U_{m_i} - 2U_{m_k} + U_{m_i+m_k} + U_{|m_i-m_k|}] =
        2[U_0 - U_{m_k} - V_0 + V_{m_k}],
\end{equation}
can be solved through standard methods of linear algebra.

@@

\section{Error estimation and mode selection}

Value selection for integer parameters is quite straightforward: try different parameter sets,
estimate error and choose best in some sense. Key process here is calculation of that estimate. One
of the possible ways is to use mean square error (\ref{lambda}) from previous section but I use
more accurate method.

Combined effect of all stages of gaussian blur approximation can be written in generic form
\begin{equation}
    R_i = \sum_{j\in Z} K_{ij}F_j,
\end{equation}
where
\begin{equation}
    K_{ij} \approx G_{i-j} = \Gamma(\sigma^2;\, i - j) =
        \frac1{\sqrt{2\pi\sigma^2}}\exp\left\{-\frac{(i - j)^2}{2\sigma^2}\right\}.
\end{equation}
Assuming that filter doesn't distort constant function (if $F_i = c$ then $R_i = c$) or
alternatively
\begin{equation}
    \sum_{j\in Z} K_{ij} = 1,
\end{equation}
and source values lie between $0$ and $1$,
\begin{equation}
    0 \le F_i \le 1,
\end{equation}
we can calculate exact upper bound of error $E^{\max}_i = \sup_F |E_i|$, where
\begin{equation}\label{E}
    E_i = \sum_{j\in Z} [K_{ij} - G_{i-j}]F_j.
\end{equation}
Due to zero error in case of zero source function
\begin{equation}
    E^{\max}_i = \sup_F |E_i| = \max\{-\inf_F E_i,\; \sup_F E_i\}.
\end{equation}
To calculate maximum positive and negative error one can choose $F_j$ in such a way that sum from
(\ref{E}) reduces to sum of only positive or negative terms,
\begin{align}
    \inf_F E_i &= \sum_{j\in Z, K_{ij} < G_{i-j}} [K_{ij} - G_{i-j}],\\
    \sup_F E_i &= \sum_{j\in Z, K_{ij} > G_{i-j}} [K_{ij} - G_{i-j}].
\end{align}
It's easy to see that
\begin{align}
    \sup_F E_i + \inf_F E_i &= \sum_{j\in Z} [K_{ij} - G_{i-j}] = 1 - 1 = 0,\\
    \sup_F E_i - \inf_F E_i &= \sum_{j\in Z} |K_{ij} - G_{i-j}|,
\end{align}
and
\begin{equation}
    E^{\max}_i = \max\{-\inf_F E_i,\; \sup_F E_i\} = \sup_F E_i =
        \frac12\sum_{j\in Z} |K_{ij} - G_{i-j}|.
\end{equation}
That's the final form of error estimation I'll use later.

Considerable experimentation shows that there is 3 types of main filter pattern worth exploring.
Those are
\begin{align}
    m^\alpha_i &= \{1, 2, 3, 4\},\\
    m^\beta_i  &= \{1, 2, 3, 5\},\\
    m^\gamma_i &= \{1, 2, 4, 6\}.
\end{align}
I call it ``1234'', ``1235'' and ``1246'' respectively. Also I consider the following types of
binomial auxiliary stage: $l = 0$ (do nothing), $l = 1$ (\ref{B2}), $l = 2$ (\ref{B4}) and $l = 3$
(\ref{B6}). Specific values of error bound is depicted at fig.~\ref{modes} for different
supplementary and main filter pairs. We can see that uniform precision is limited by error of 0.12\%
for $\sigma^2 \approx 6.693$. Taking that value into consideration and also the fact that use of
lower values of $l$ is preferable (better performance) I ended up with the following strategy.
\begin{itemize}
    \item If $\sigma^2 < 1.9$ then use single filter ``1234'' in direct mode---with coefficients
        from (\ref{gamma}). Don't perform any rescaling ($n = 0$) and don't use any additional
        filters ($l = 0$).
    \item Otherwise if $1.9 \le \sigma^2 < 6.693$ then skip rescaling ($n = 0$) but use nontrivial
        auxiliary filter.
        \begin{itemize}
            \item For $1.9 \le \sigma^2 < 2.8$   use filter ``1234'' with $\hat B^2$ ($l = 1$).
            \item For $2.8 \le \sigma^2 < 4.4$   use filter ``1235'' with $\hat B^4$ ($l = 2$).
            \item For $4.4 \le \sigma^2 < 6.693$ use filter ``1246'' with $\hat B^6$ ($l = 3$).
        \end{itemize}
    \item Else ($\sigma^2 \ge 6.693$) calculate number of shrinking/expanding stages as
        \begin{equation}
            n = 2 + \left\lfloor\log_4\frac{\sigma^2 + 0.7}{26.5}\right\rfloor,
        \end{equation}
        and scaled dispersion
        \begin{equation}
            \tilde\sigma^2 = \sigma^2\cdot4^{-n}.
        \end{equation}
        \begin{itemize}
            \item If $\tilde\sigma^2 < 3.15 - 1.5\cdot4^{-n}$ then
                use single filter ``1234'' ($l = 0$).
            \item Else if $\tilde\sigma^2 < 5.3 - 5.2\cdot4^{-n}$ then
                use filter ``1235'' with $\hat B^2$ ($l = 1$).
            \item Otherwise use filter ``1246'' with $\hat B^4$ ($l = 2$).
        \end{itemize}
\end{itemize}
There is a possibility that such strategy can result in negative main filter coefficients $C_i$
(\ref{main}).  It's undesirable due to probability of overflow of initial $[0, 1]$ range. So I
truncate such coefficients to zero. That increases error in certain areas but do not degrades global
precision. The final error bound achieved by the full algorithm is displayed in fig.~\ref{error}.

\pgfplotstableread{error.table}\errortable
\begin{figure}[t]
    \centering
    \begin{tikzpicture}
        \begin{semilogxaxis} [ title = $E_{\max}(\sigma^2)$,
                width = 18cm, height = 6cm, scaled ticks = false,
                xmin = 1, xmax = 10000, xminorgrids = true, xmajorgrids = true,
                ymin = 0, ymax = 0.003, yminorgrids = true, ymajorgrids = true,
                yticklabel style = { overlay, /pgf/number format/.cd, fixed, fixed zerofill, precision = 3 },
                ytick = { 0, 0.001, 0.002, 0.003 }, minor y tick num = 1 ]
            \addlegendentry{$1234, l = 0$}
            \addplot [ sharp plot, red ]
                table[x = r2, y = l0-1234] {\errortable};
            \addlegendentry{$1234, l = 1$}
            \addplot [ sharp plot, densely dotted, red ]
                table[x = r2, y = l1-1234] {\errortable};
            \addlegendentry{$1235, l = 1$}
            \addplot [ sharp plot, green ]
                table[x = r2, y = l1-1235] {\errortable};
            \addlegendentry{$1235, l = 2$}
            \addplot [ sharp plot, densely dotted, green ]
                table[x = r2, y = l2-1235] {\errortable};
            \addlegendentry{$1246, l = 2$}
            \addplot [ sharp plot, blue ]
                table[x = r2, y = l2-1246] {\errortable};
            \addlegendentry{$1246, l = 3$}
            \addplot [ sharp plot, densely dotted, blue ]
                table[x = r2, y = l3-1246] {\errortable};
        \end{semilogxaxis}
    \end{tikzpicture}
    \caption{Error bound of different sets of filters.}
    \label{modes}
\end{figure}
\begin{figure}[t]
    \centering
    \begin{tikzpicture}
        \begin{semilogxaxis} [ title = $E_{\max}(\sigma^2)$,
                width = 18cm, height = 6cm, scaled ticks = false,
                xmin = 1, xmax = 10000, xminorgrids = true, xmajorgrids = true,
                ymin = 0, ymax = 0.0015, yminorgrids = true, ymajorgrids = true,
                yticklabel style = { overlay, /pgf/number format/.cd, fixed, fixed zerofill, precision = 4 },
                ytick = { 0, 0.0005, 0.001, 0.0015 }, minor y tick num = 4 ]
            \addplot [ sharp plot, line width = 1pt, red ]
                table[x = r2, y = final] {\errortable};
        \end{semilogxaxis}
    \end{tikzpicture}
    \caption{Error bound of the full algorithm.}
    \label{error}
\end{figure}

Previous plots show only range $\sigma^2 > 1$ but there is another region that requires special
attention. Problem of small blur radius is discussed in the next section.
